The \proglang{R} Package \pkg{forestinventory}: Design-Based Global and Small Area Estimations for Multi-Phase Forest Inventories

### 2017-10-02 rnw/Threephase_Estimators and their Application.Rnw
update Andreas 02.10.2017

The the vector of regression coefficients for the \textbf{\textit{reduced model}} using the \textit{reduced set of explanatory variables} $\pmb{Z}^{(0)}(x)$ available at all sample locations $x \in s_0$ is derived by solving the normal equation

\begin{eqnarray}\label{eq:normequ_redmod}
\hat{\pmb{\alpha}}_{s_2} &=& \Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{Z}^{(0)}(x)
\end{eqnarray}

and likewise the vector of regression coefficients for the \textbf{\textit{full model}} using the \textit{full set of explanatory variables} $\pmb{Z}^t(x)=(\pmb{Z}^{(0)t}(x),\pmb{Z}^{(1)t}(x))$ available only at a subset of all sample locations $x \in s_1 \subset s_0$:

\begin{eqnarray}\label{eq:normequ_redmod}
\hat{\pmb{\beta}}_{{s_2}}&=&\Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{Z}(x)\pmb{Z}^t(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{Z}(x)
\end{eqnarray}

The variance-covariance matrices of the regression coefficients are thus defined as

\begin{subequations}\label{eq:covar3p}
\begin{align}
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}_{0}^2(x)\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x)\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x) \Big)^{-1} \label{eq:covar_alpha} \\
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}(x)\pmb{Z}^t(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}^2(x)\pmb{Z}(x)\pmb{Z}^t(x)\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}(x)\pmb{Z}^t(x) \Big)^{-1} \label{eq:covar_beta}
\end{align}
\end{subequations}

with the full model residuals ${R}(x)=Y(x)-\pmb{Z}^t(x)\hat{\beta}_{s_2}$ and the reduced model residuals ${R}_{0}(x)=Y(x)-\pmb{Z}^{(0)t}\hat{\alpha}_{s_2}$ derived under the reduced and the full regression model respectively.

Likewise the two-phase estimation (section \ref{sec:twophase_and_appl}), also the three-phase estimation can be applied under \textit{non-exhaustive} and \textit{exhaustive} derivation of the explanatory variables at the $s_0$ phase.

In the \textbf{non-exhaustive case}, we use the \textit{estimated} means of $\pmb{Z}^{(0)}(x)$ over both the $s_0$ and $s_1$ phase, as well as the \textit{estimated} mean of $\pmb{Z}(x)$ over the $s_1$ phase (equation \ref{meanvalues3pnex}). If the explanatory variables used in the reduced model are exhaustively derived (\textbf{exhaustive case}), the estimated mean $\hat{\bar{\pmb{Z}}}^{(0)}_0$ is simply replaced by the exact mean $\bar{\pmb{Z}}^{(0)}_0$. Note that in case of applied boundary adjustment (section \ref{sec:twophase_and_appl}), the simple mean is again replaced by the weighted mean analogous to equation \ref{eq:wmeanZ}.

@@

The internal model case again insures the zero mean residual property over $F$, and the \textbf{point estimates} are thus calculated as

Equations \ref{eq:reg3p_nex} and \ref{eq:reg3p_ex} nicely demonstrate the philosophy of the three-phase estimation approach. Without the first term, the equations exactly equal the one of two-phase estimator where the estimated means in $\hat{\bar{\pmb{Z}}}_1$ are based on the $s_1$ sample and are transformed into predictions $\hat{Y}(x)$ of the local density $Y(x)$. The difference however is that we can now derive a more precise mean $\hat{\bar{\pmb{Z}}}^{(0)}_0$ (or even the exact mean $\bar{\pmb{Z}}^{(0)}_0$) over $s_0$ for a subset $\pmb{Z}^{(0)}(x) \in \pmb{Z}(x)$. The first term of equation \ref{eq:reg3p_nex} and \ref{eq:reg3p_ex} can thus be interpreted as a correction term that transforms the difference to the more precise / exact mean of $\pmb{Z}^{(0)}(x)$ into a residual of $Y(x)$ and adds it to the second term, thus making the point estimate more precise.

The \textbf{design-based variance} is then calculated as

@@

with the variance-covariance matrix $\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}$:

\begin{equation}\label{estvarcovaux3p}
\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}=
\frac{1}{n_{0}(n_{0}-1)}\sum_{x\in{s_{0}}}
(\pmb{Z}^{(0)}(x)-\hat{\bar{\pmb{Z}}}^{(0)}_{0})(\pmb{Z}^{(0)}(x)-\hat{\bar{\pmb{Z}}}^{(0)}_{0})^t
\end{equation}

We see that equation \ref{eq:dbvar_reg3p_ex} (exhaustive case) and \ref{eq:dbvar_reg3p_nex} (non-exhaustive case) differ by the term $\hat{\pmb{\alpha}}_{s_2}^t\hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}\hat{\pmb{\alpha}}_2$ which is exactly the variance of the prediction $\hat{Y}_{0}(x)$ of the reduced model in the $s_0$ phase. This term can be made very small by chosing a large sample size $n_0$ for $s_0$ \citep{mandallaz2013techb}.

The package also gives the \textbf{external variance} defined as

@@

@@

Note that the \code{grisons} inventory data set originally corresponds to a classical two-phase sample design, i.e. the full set of explanatory variables is available at all sample locations in the inventory area. To demonstrate the three-phase estimation procedure, we provide an artificial three-phase scenario by recoding the phase indicators in the data set (column \code{phase_id_3p}). The values for indicating the phases (\code{0,1,2}) follow the terminology used in this article ($s_0$, $s_1$, $s_2$).

We can use the \code{table()} function to see the sample sizes of all phases for the three-phase scenario:

\begin{small}
<<>>=
table(grisons$phase_id_3p)
@
\end{small}

We see that in the three-phase example, we now have 40 terrestrial sample locations providing the timber volume per hectare (\code{tvol}). The large phase $s_0$ comprises $178 + 88 + 40 = 306$ sample locations, since $s_2 \subset s_1 \subset s_0$. We now assume that at sample locations $x \in s_0$, only the mean canopy height (\code{mean}) is available, whereas at sample locations $x \in s_1$ we additionally have the explanatory variables \code{stddev}, \code{max} and \code{q75} at hand. We can now define the reduced and full regression model formulas that we will use in the \code{threephase()} function. Note that the models are nested.

@@

Note that compared to the \code{twophase()}-function, we now have to specify two regression models, i.e. the nested reduced (\code{formula.s0}) and full (\code{formula.s1}) regression model. In addition, we also have to specify the indication of the $s_1$ phase (\code{s1.id}) in the argument \code{phase_id}. The global three-phase estimation can thus be applied by

@@

The \code{summary()} of a \code{threephase()}-function recalls both regression model formulas and also gives the coefficients of determination $R^2$ for both the reduced (\code{r.squared_reduced}) and the full (\code{r.squared_full}) regression model. The set of sample sizes have also been extented by the $s_0$ phase (\code{n0}). In order to compare the results to the respective two-phase estimation, we assume that we have the mean canopy height as the only explanatory variable at all 306 sample locations:

\begin{small}
<<>>=
reg2p_nex<- twophase(formula = formula.rm, data = grisons,
                       phase_id = list(phase.col="phase_id_2p", terrgrid.id = 2),
                       boundary_weights = "boundary_weights")
@
\end{small}

\begin{small}
<<>>=
reg2p_nex$estimation
@
\end{small}

Comparing the $R^2$s of the three and two-phase result, it becomes obvious that the incorporation of the additional explanatory variables \code{stddev}, \code{max} and \code{q75} led to a 20\% increase in explained variance in the regression model. While the more precise regression model was only applied at 40\% of all 306 sample locations (\textit{sampling fraction}), it was possible to considerably reduce the two-phase variance and thus achieve a higher estimation precision.

@@

The 3 small area estimators under three-phase sampling provided by the package extend the two-phase small area estimator of section \ref{sec:twophase_sae} to the three-phase case. Note that the general principle stays the same, i.e. all information used is restricted to the small area $G$, with exception of retrieving the regession coefficients of the reduced and full model and their variance-covariance matrices.

We first define the estimated means of the explanatory variables within the small area $G$:

\begin{equation}\label{meanvalues3psaenext}
\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}=\frac{1}{n_{0,G}}\sum_{x\in{s_{0,G}}} \pmb{Z}^{(0)}(x), \quad \hat{\bar{\pmb{Z}}}^{(0)}_{1,G}=\frac{1}{n_{1,G}}\sum_{x\in{s}_{1,G}}\pmb{Z}^{(0)}(x) ,
\quad \hat{\bar{\pmb{Z}}}_{1,G}=\frac{1}{n_{1,G}}\sum_{x\in{s}_{1,G}}\pmb{Z}(x)
\end{equation}

@@

The \textbf{small} (\textit{small}) and \textbf{pseudo small area estimator} (\textit{psmall}) follow the same principle as in the two-phase case, i.e. applying the globally derived reduced and full regression model to the small area means of the explanatory variables, and then correcting for a potential model bias by adding the residual mean of the full model, which is not zero in general, to the point estimate (equations \ref{eq:pe_3p_small} and \ref{eq:pe_3p_psmall}). Note that the difference between the mean $\hat{\bar{\pmb{Z}}}^{(0)}_{1,G}$ derived in the $s_1$ phase and the more precise mean $\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}$ or exact mean $\bar{\pmb{Z}}^{(0)}_{0,G}$ derived in the $s_0$ phase is again considered as a residual term likewise in the global estimation (equation \ref{eq:reg3p_ex} and \ref{eq:reg3p_nex}).

The \textbf{point estimates} are unbiased and given by:

\begin{subequations}\label{eq:pest_3p_small_psmall}
\begin{align}
\hat{Y}_{G,small,3p}&=(\bar{\pmb{Z}}^{(0)}_{0,G}-\hat{\bar{\pmb{Z}}}^{(0)}_{1,G})^t\hat{\pmb{\alpha}}_{s_2} +
\hat{\bar{\pmb{Z}}}^t_{1,G}\hat{\pmb{\beta}}_{s_2}+\frac{1}{n_{2,G}}\bar{\hat{R}}_{G}(x) \label{eq:pe_3p_small} \\
\hat{Y}_{G,psmall,3p}&=(\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}-\hat{\bar{\pmb{Z}}}^{(0)}_{1,G})^t\hat{\pmb{\alpha}}_{s_2} +
\hat{\bar{\pmb{Z}}}^t_{1,G}\hat{\pmb{\beta}}_{s_2}+\frac{1}{n_{2,G}}\bar{\hat{R}}_{G}(x) \label{eq:pe_3p_psmall}
\end{align}
\end{subequations}

with ${R}_{G}(x)=Y_G(x)-\pmb{Z}_G^t(x)\hat{\beta}_{s_2}$ being the full model residuals in small area $G$. The \textbf{design-based variance} is then calculated as

@@

\begin{subequations}\label{eq:pest_3p_synth_psynth}
\begin{align}
\hat{Y}_{G,synth,3p}&=(\bar{\pmb{Z}}^{(0)}_{0,G}-\hat{\bar{\pmb{Z}}}^{(0)}_{1,G})^t\hat{\pmb{\alpha}}_2 +
\hat{\bar{\pmb{Z}}}^t_{1,G}\hat{\pmb{\beta}}_{s_2} \label{eq:pe_3p_synth} \\
\hat{Y}_{G,psynth,3p}&=(\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}-\hat{\bar{\pmb{Z}}}^{(0)}_{1,G})^t\hat{\pmb{\alpha}}_2 +
\hat{\bar{\pmb{Z}}}^t_{1,G}\hat{\pmb{\beta}}_{s_2}\label{eq:pe_3p_psynth}
\end{align}
\end{subequations}

\begin{subequations}\label{eq:var_3p_synth_psynth}
\begin{align}
\hat{\var}(\hat{Y}_{G,synth,3p})& =\frac{n_2}{n_1}\bar{\pmb{Z}}^{(0)t}_{0,G}\hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}
\bar{\pmb{Z}}^{(0)}_{0,G}+(1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}_{1,G}^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}
\hat{\bar{\pmb{Z}}}_{1,G} \label{eq:var_3p_reg_synth}\\
\hat{\var}(\hat{Y}_{G,psynth,3p})& = \hat{\pmb{\alpha}}_2^t\hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}}\hat{\pmb{\alpha}}_2 +
\frac{n_2}{n_1}\hat{\bar{\pmb{Z}}}^{(0)t}_{0,G}\hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}
\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}+(1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}_{1,G}^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}
\hat{\bar{\pmb{Z}}}_{1,G} \label{eq:var_3p_reg_psynth}
\end{align}
\end{subequations}

@@

The \textbf{extended synthetic} (\textit{extsynth}) and \textbf{extended pseudo synthetic estimator} (\textit{extpsynth}) insures that the residuals of the full model over both the entire inventory area $F$ and the small area $G$ are zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This is again realized by extending the vector of explanatory variables by a binary categorical indicator variable $I_G(x)$ which takes the value 1 if the observation lies inside the small area $G$ and is otherwise set to 0. The extended auxiliary vector is thus defined as $\pmb{\mathbb{Z}}^t(x)=(\pmb{\mathbb{Z}}^{(0)t}(x),\pmb{\mathbb{Z}}^{t}(x))$, where $\pmb{\mathbb{Z}}^{(0)t}(x)=(\pmb{Z}^{(1)t}(x), I_G^t(x))$ and $\pmb{\mathbb{Z}}^{t}(x)=\pmb{Z}^{t}(x)$. This practically means that we have to specify the binary indicator variable for \textit{all} observations in the data frame that is passed to the package-functions.

We first recalculate the regression coefficients $\hat{\pmb{\gamma}}_{s_2}$ and $\hat{\pmb{\theta}}_{s_2}$, and their variance-covarianve matrices $\hat{\pmb{\Sigma}}_{\hat{\pmb{\gamma}}_{s_2}}$ and $\hat{\pmb{\Sigma}}_{\hat{\pmb{\theta}}_{s_2}}$ under the extended auxiliary vector as

\begin{subequations}\label{eq:normequ_redmod_ext}
\begin{align}
\hat{\pmb{\gamma}}_{s_2} &= \Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{\mathbb{Z}}^{(0)}(x)\pmb{\mathbb{Z}}^{(0)t}(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{\mathbb{Z}}^{(0)}(x) \\
\hat{\pmb{\theta}}_{{s_2}}&=\Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{\mathbb{Z}}(x)
\end{align}
\end{subequations}

and

>>>>

Solving the sample based normal equations, the vector of regression coefficients $\hat{\pmb{\alpha}}_{s_2}$ for the \textbf{\textit{reduced model}}, i.e. using the \textit{reduced set of explanatory variables} $\pmb{Z}^{(0)}(x)$ available at $x \in s_0$, and likewise the vector of regression coefficients $\hat{\pmb{\beta}}_{s_2}$ for the \textbf{\textit{full model}}, i.e. using the \textit{full set of explanatory variables} $\pmb{Z}^t(x)=(\pmb{Z}^{(0)t}(x),\pmb{Z}^{(1)t}(x))$ available only at a subset $x \in s_1 \subset s_0$, are derived as

\begin{subequations}\label{eq:normequ_3p}
\begin{align}
\hat{\pmb{\alpha}}_{s_2}&=\Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{Z}^{(0)}(x)  \label{eq:normequ_redmod} \\
\hat{\pmb{\beta}}_{s_2}&=\Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{Z}(x)\pmb{Z}^t(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{Z}(x) \label{eq:normequ_fullmod}
\end{align}
\end{subequations}

The package allows for the calculation of \textbf{point estimates} under exhaustive and non-exhaustive use of the auxiliary information in the $s_0$ phase. Staying in the internal model approach again insures the zero mean residual property over $F$.

In the \textit{non-exhaustive case}, the \textit{estimated} means of $\pmb{Z}^{(0)}(x)$ over both the $s_0$ and $s_1$ phase, as well as the \textit{estimated} mean of $\pmb{Z}(x)$ over the $s_1$ phase are calculated according to \ref{meanvalues3pnex}. If the exact mean over $s_0$ is known, the estimated mean $\hat{\bar{\pmb{Z}}}^{(0)}_0$ can simply be replaced by the exact mean $\bar{\pmb{Z}}^{(0)}_0$. Note that in case of applied boundary adjustment (section \ref{sec:twophase_and_appl}), the simple mean is again replaced by the weighted mean analogous to equation \ref{eq:wmeanZ}.

@@

An intuitive interpretation of \ref{eq:reg3p_ex} and \ref{eq:reg3p_nex} is a follows: Likewise in the twophase approach, the estimated means in $\hat{\bar{\pmb{Z}}}_1$ based on the $s_1$ sample are transformed into predictions $\hat{Y}(x)$ of the local density $Y(x)$. The difference between the estimated means $\hat{\bar{\pmb{Z}}}^{(0)}_1$ and their more precise or even exact counterparts $\hat{\bar{\pmb{Z}}}^{(0)}_0$ and $\bar{\pmb{Z}}^{(0)}_0$ based on the larger or infinite $s_0$ sample is then added as a residual term that expresses the lack of model precision due to varying sample sizes behind the estimated auxilary means.

The package again provides the g-weight and external variances. The \textbf{g-weight variance} formulation is

@@

with the variance-covariance matrix of $\hat{\bar{\pmb{Z}}}^{(0)}_0$ and the variance-covariance matrices of the regression coefficients $\hat{\pmb{\alpha}}_{s_2}$ and $\hat{\pmb{\beta}}_{s_2}$:

\begin{subequations}\label{eq:covar3p}
\begin{align}
  \hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}&=
  \frac{1}{n_{0}(n_{0}-1)}\sum_{x\in{s_{0}}}
  (\pmb{Z}^{(0)}(x)-\hat{\bar{\pmb{Z}}}^{(0)}_{0})(\pmb{Z}^{(0)}(x)-\hat{\bar{\pmb{Z}}}^{(0)}_{0})^t \\
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}&=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}_{0}^2(x)\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x)\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x) \Big)^{-1} \label{eq:covar_alpha} \\
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}&=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}(x)\pmb{Z}^t(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}^2(x)\pmb{Z}(x)\pmb{Z}^t(x)\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}(x)\pmb{Z}^t(x) \Big)^{-1} \label{eq:covar_beta}
\end{align}
\end{subequations}

Note that ${R}(x)=Y(x)-\pmb{Z}^t(x)\hat{\beta}_{s_2}$ denotes the residuals of the full model, whereas ${R}_{0}(x)=Y(x)-\pmb{Z}^{(0)t}\hat{\alpha}_{s_2}$ denotes the residuals of the reduced model. The \textbf{external variance} form under linear regression models is defined as

@@

% We see that equation \ref{eq:dbvar_reg3p_ex} (exhaustive case) and \ref{eq:dbvar_reg3p_nex} (non-exhaustive case) differ by the term $\hat{\pmb{\alpha}}_{s_2}^t\hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}\hat{\pmb{\alpha}}_2$ which is exactly the variance of the prediction $\hat{Y}_{0}(x)$ of the reduced model in the $s_0$ phase. This term can be made very small by chosing a large sample size $n_0$ for $s_0$ \citep{mandallaz2013techb}.

@@

In order to demonstrate the three-phase estimators in the package, we created an artificial three-phase scenario by recoding the phase indicators in the \code{grisons} data set (column \code{phase_id_3p}) according to the terminology used in this article (\code{0} for $s_0$, \code{1} for $s_1$, \code{2} for $s_2$). We now assume that the mean canopy height (\code{mean}) is available at all 306 sample locations $x \in s_0$, whereas we have the explanatory variables \code{stddev}, \code{max} and \code{q75} only at 128 subsamples $s_1$ of $s_0$. At 40 further subsamples $s_2$ we have the observations $Y(x)$ from the field inventory. Based on this setup, we can now define the reduced and full regression model formulas to be used in the \code{threephase()} function (note that the models are nested).

@@

Compared to the \code{twophase()}-function, we now have to specify two regression models, i.e. the nested reduced (\code{formula.s0}) and full (\code{formula.s1}) regression model. In addition, we also have to specify the indication of the $s_1$ phase (\code{s1.id}) in the argument \code{phase_id}. The global three-phase estimation can thus be applied by

@@

The \code{summary()} of a \code{threephase()}-function now recalls \textit{both} regression model formulas and also gives the coefficients of determination $R^2$ for both the reduced (\code{r.squared_reduced}) and the full (\code{r.squared_full}) regression model. We are told that including \code{stddev}, \code{max} and \code{q75} besides \code{mean} yields a 20 \% improvement in R$^2$. If we used \code{mean} as the only auxiliary variable at all sample locations (two-phase approach), we would see that the three-phase extention led to a considerable reduction in variance.

@@

The three two-phase small area estimators described in section \ref{sec:twophase_sae} can also be extended to the three-phase scenario. The general principle thereby stays the same, i.e. the regession coefficients of the reduced and full model and their variance-covariance matrices are calculated on the entire $s_2$ sample according to equations \ref{eq:normequ_redmod}, \ref{eq:normequ_fullmod}, \ref{eq:covar_alpha} and \ref{eq:covar_beta}, and are subsequently used to make predictions for sample locations restricted to small area $G$.

% We first define the estimated means of the explanatory variables within the small area $G$:
% 
% \begin{equation}\label{meanvalues3psaenext}
% \hat{\bar{\pmb{Z}}}^{(0)}_{0,G}=\frac{1}{n_{0,G}}\sum_{x\in{s_{0,G}}} \pmb{Z}^{(0)}(x), \quad \hat{\bar{\pmb{Z}}}^{(0)}_{1,G}=\frac{1}{n_{1,G}}\sum_{x\in{s}_{1,G}}\pmb{Z}^{(0)}(x) ,
% \quad \hat{\bar{\pmb{Z}}}_{1,G}=\frac{1}{n_{1,G}}\sum_{x\in{s}_{1,G}}\pmb{Z}(x)
% \end{equation}

@@

The unbiased \textbf{point estimates} of the \textbf{small} (\textit{small}) and \textbf{pseudo small area estimator} (\textit{psmall}) are calculated by applying the globally derived reduced and full regression model coefficients to the small area means of the explanatory variables, and then correcting for a potential model bias in $G$ by adding the small area mean of the full model residuals, i.e. ${R}_{G}(x)=Y_G(x)-\pmb{Z}_G^t(x)\hat{\beta}_{s_2}$, to the point estimate. The difference between the mean $\hat{\bar{\pmb{Z}}}^{(0)}_{1,G}$ and the more precise or exact mean $\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}$ and $\bar{\pmb{Z}}^{(0)}_{0,G}$ is again considered as a residual term likewise in the global estimation (equation \ref{eq:reg3p_ex} and \ref{eq:reg3p_nex}).

\begin{subequations}\label{eq:pest_3p_small_psmall}
\begin{align}
\hat{Y}_{G,small,3p}&=(\bar{\pmb{Z}}^{(0)}_{0,G}-\hat{\bar{\pmb{Z}}}^{(0)}_{1,G})^t\hat{\pmb{\alpha}}_{s_2} +
\hat{\bar{\pmb{Z}}}^t_{1,G}\hat{\pmb{\beta}}_{s_2}+\frac{1}{n_{2,G}}\hat{R}_{G}(x) \label{eq:pe_3p_small} \\
\hat{Y}_{G,psmall,3p}&=(\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}-\hat{\bar{\pmb{Z}}}^{(0)}_{1,G})^t\hat{\pmb{\alpha}}_{s_2} +
\hat{\bar{\pmb{Z}}}^t_{1,G}\hat{\pmb{\beta}}_{s_2}+\frac{1}{n_{2,G}}\hat{R}_{G}(x) \label{eq:pe_3p_psmall}
\end{align}
\end{subequations}

The \textbf{g-weight variance} is then calculated as

@@

\begin{subequations}\label{eq:3p_synth_psynth}
\begin{align}
\hat{Y}_{G,synth,3p}&=(\bar{\pmb{Z}}^{(0)}_{0,G}-\hat{\bar{\pmb{Z}}}^{(0)}_{1,G})^t\hat{\pmb{\alpha}}_2 +
\hat{\bar{\pmb{Z}}}^t_{1,G}\hat{\pmb{\beta}}_{s_2} \label{eq:pe_3p_synth} \\
\hat{Y}_{G,psynth,3p}&=(\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}-\hat{\bar{\pmb{Z}}}^{(0)}_{1,G})^t\hat{\pmb{\alpha}}_2 +
\hat{\bar{\pmb{Z}}}^t_{1,G}\hat{\pmb{\beta}}_{s_2}\label{eq:pe_3p_psynth} \\
\hat{\var}(\hat{Y}_{G,synth,3p})& =\frac{n_2}{n_1}\bar{\pmb{Z}}^{(0)t}_{0,G}\hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}
\bar{\pmb{Z}}^{(0)}_{0,G}+(1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}_{1,G}^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}
\hat{\bar{\pmb{Z}}}_{1,G} \label{eq:var_3p_reg_synth}\\
\hat{\var}(\hat{Y}_{G,psynth,3p})& = \hat{\pmb{\alpha}}_2^t\hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}}\hat{\pmb{\alpha}}_2 +
\frac{n_2}{n_1}\hat{\bar{\pmb{Z}}}^{(0)t}_{0,G}\hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}
\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}+(1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}_{1,G}^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}
\hat{\bar{\pmb{Z}}}_{1,G} \label{eq:var_3p_reg_psynth}
\end{align}
\end{subequations}

% \begin{subequations}\label{eq:var_3p_synth_psynth}
% \begin{align}
% \hat{\var}(\hat{Y}_{G,synth,3p})& =\frac{n_2}{n_1}\bar{\pmb{Z}}^{(0)t}_{0,G}\hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}
% \bar{\pmb{Z}}^{(0)}_{0,G}+(1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}_{1,G}^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}
% \hat{\bar{\pmb{Z}}}_{1,G} \label{eq:var_3p_reg_synth}\\
% \hat{\var}(\hat{Y}_{G,psynth,3p})& = \hat{\pmb{\alpha}}_2^t\hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_{0,G}}\hat{\pmb{\alpha}}_2 +
% \frac{n_2}{n_1}\hat{\bar{\pmb{Z}}}^{(0)t}_{0,G}\hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}
% \hat{\bar{\pmb{Z}}}^{(0)}_{0,G}+(1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}_{1,G}^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}
% \hat{\bar{\pmb{Z}}}_{1,G} \label{eq:var_3p_reg_psynth}
% \end{align}
% \end{subequations}

@@

The \textbf{extended synthetic} (\textit{extsynth}) and \textbf{extended pseudo synthetic estimator} (\textit{extpsynth}) insures that the residuals of the full model over both the entire inventory area $F$ and the small area $G$ are zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This is again realized by extending the vector of explanatory variables by a binary categorical indicator variable $I_G(x)$ which takes the value 1 if the observation lies inside the small area $G$ and is otherwise set to 0. The extended auxiliary vector is thus defined as $\pmb{\mathbb{Z}}^t(x)=(\pmb{\mathbb{Z}}^{(0)t}(x),\pmb{\mathbb{Z}}^{t}(x))$, where $\pmb{\mathbb{Z}}^{(0)t}(x)=(\pmb{Z}^{(1)t}(x), I_G^t(x))$ and $\pmb{\mathbb{Z}}^{t}(x)=\pmb{Z}^{t}(x)$. This practically means that we have to specify the binary indicator variable for \textit{all} observations in the data frame that is passed to the package-functions. The regression coefficients $\hat{\pmb{\gamma}}_{s_2}$ and $\hat{\pmb{\theta}}_{s_2}$, and their variance-covarianve matrices $\hat{\pmb{\Sigma}}_{\hat{\pmb{\gamma}}_{s_2}}$ and $\hat{\pmb{\Sigma}}_{\hat{\pmb{\theta}}_{s_2}}$ under the extended auxiliary vector are thus recalculated as

\begin{subequations}\label{eq:normequ_redmod_ext}
\begin{align}
\hat{\pmb{\gamma}}_{s_2} &= \Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{\mathbb{Z}}^{(0)}(x)\pmb{\mathbb{Z}}^{(0)t}(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{\mathbb{Z}}^{(0)}(x) \\
\hat{\pmb{\theta}}_{{s_2}}&=\Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{\mathbb{Z}}(x) \\
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\gamma}}_{s_2}}&=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}^{(0)}(x)\pmb{\mathbb{Z}}^{(0)t}(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}_{0}^2(x)\pmb{\mathbb{Z}}^{(0)}(x)\pmb{\mathbb{Z}}^{(0)t}(x)\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}^{(0)}(x)\pmb{\mathbb{Z}}^{(0)t}(x) \Big)^{-1} \\
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\theta}}_{s_2}}&=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}^2(x)\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x)\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1}
\end{align}
\end{subequations}

With the auxliary means and their variance-covariance matrices following the same notational change, the \textbf{point estimate} and \textbf{g-weight variance} are derived as
